Factor Analysis on Report Citations, using a Combined Latent and logistic Model

### 2019-05-13 Manuscript for submission/manuscript.tex
latex works

\author[3]{Eric Heim}

\author[3]{Timothy Van Slyke}

\author[3]{Lee Seversky}

@@

\address[3]{\orgdiv{Org Division}, \orgname{Org Name}, \orgaddress{\state{State name}, \country{Country name}}}

@@

\section{Introduction}\label{sec:intro}
We study a citation network, where each node (i.e., item) can be a technical report or a publication. A node may cite another node. Associated with a pair of nodes $i$ and $j$, we denote a binary random variable $X_{ij}$, where $1 \le i,j \le n$ and $n$ is the total number of nodes. We have $X_{ij}=1$ if and only if either node $i$ cites node $j$ or vice versa; otherwise $X_{ij}=0$. For each node $i$, we assume that there is an associated binary vector $f_{i} \in \mathbb{R}^K$, such that the $k$th entry of $f_{i}$, $f_{ik} = 1$, if and only if node $i$ is related to topic (i.e., factor) $k, 1\le k \le K$. Here $K$ is the total number of underlying topics (i.e., factors, or trends). We assume a logistic model for $X_{ij}$'s: for $1\le i,j \le n$,

\begin{equation} 
\label{eq:logistic01}
\mathbb{P}(X_{ij}=1) = \frac{e^{\alpha + f_i^T D f_j }}{1 + e^{\alpha + f_i^T D f_j }},
\end{equation}
here $\alpha \in \mathbb{R}$ is a parameter and matrix $D \in \mathbb{R}^{K \times K}$ is a diagonal matrix: $D = \mbox{diag}\{d_{1},d_{2},\ldots,d_{K}\}$. We assume $d_{ii} > 0$ for $1 \le i \le K$.
Another way to put \eqref{eq:logistic01} is
\begin{equation}
\label{eq:logistic02}
\mathbb{P}(X_{ij}=1) = \frac{\exp\left\{\alpha + \sum_{k=1}^K f_{ik}f_{jk}d_{k} \right\}}{1 + \exp\left\{\alpha + \sum_{k=1}^K f_{ik}f_{jk}d_{k} \right\}},
\end{equation}
A justification of the above model is that when both node $i$ and node $j$ are related to topic $k$, they have a higher chance to cite one way or the other. We have assumed a common strengthen coefficient $d_k$ ($1\le k \le K$) for factor $k$, despite different nodes. We denote a matrix $F = \{f_1, f_2, \ldots, f_n\} \in \mathbb{R}^{K \times n}$. Each column $i$ in matrix $F$ contains the factor loadings associated with the node $i$ ($1\le i \le n$). Given the diagonal matrix $D$ and the factor loading matrix $F$, we assume that $X_{ij}$'s are independent; therefore we have the total conditional probability function as follows: 
\begin{equation}
\label{eq:logistic03}
\mathbb{P}(\{X_{ij}, 1\le i,j \le n\})
= \prod_{1\le i<j \le n} \mathbb{P}(X_{ij})
= \prod_{1\le i<j \le n}  \frac{e^{X_{ij}(\alpha + f_i^T D f_j) }}{1 + e^{\alpha + f_i^T D f_j }},
\end{equation}
where $\mathbb{P}(X_{ij})$ is given in \eqref{eq:logistic02}.
The last equation holds because $X_{ij}$ only takes binary (i.e., $0$ or $1$) values.
Recall that the dot product of two matrices with same dimensionality, $A,B\in \mathbb{R}^{a \times b}$, is defined as $A\bullet B=\mbox{trace}(A^T B) = \sum_{i=1}^a\sum_{j=1}^b a_{ij}b_{ij}$.
The above \eqref{eq:logistic03} can be further rewritten as
\begin{equation}
\label{eq:logistic04}
\mathbb{P}(\{X_{ij}, 1\le i,j \le n\})
= \frac{\exp(\alpha \sum_{1\le i< j\le n}X_{ij} +\frac{1}{2} X \bullet (F^T D F))}{\prod_{1\le i<j \le n}  1 + e^{f_i^T D f_j }},
\end{equation}
where we assume $X_{ii}=0$ for all $i$ ($1\le i \le n$) and $X_{ij} = X_{ji}$ for all $i$ and $j$ ($1\le i,j \le n$), i.e., the matrix $X$ is symmetric.
The above delivers a factor analysis model. Various linear and nonlinear latent variable models have been studied extensively in the literature (e.g.,\cite{joreskog1969general, mcdonald2014factor,lord2008statistical, rasch1980probabilistic, harman1960modern, joreskog1970general}).

@@

We now present each of the three steps of the ADMM algorithm and show that the proximal
operators $\mathbf{P}_{\lambda,f}$ and $\mathbf{P}_{\lambda,g}$ are easy to evaluate.
Let
$$
x^m = (x^m_\alpha, x^m_M, x^m_L, x^m_S), \quad
z^m = (z^m_\alpha, z^m_M, z^m_L, z^m_S), \quad
u^m = (u^m_\alpha, u^m_M, u^m_L, u^m_S).
$$
Step 1. We solve $x^{m+1} = \mathbf{P}_{\lambda,f}(z^m - u^m)$.
Due to the special structure of $f(\cdot)$,
$x^{m+1}_\alpha, x^{m+1}_M, x^{m+1}_L$, and $x^{m+1}_S$
can be updated separately. More precisely, we have
\begin{eqnarray}
x^{m+1}_\alpha, x^{m+1}_M &=& \mbox{arg min}_{\alpha, M} \quad
-\frac{\alpha}{n} \sum_{1\le i< j\le n}X_{ij}
-\frac{1}{2n} X \bullet M
+ \frac{1}{n} \sum_{1\le i<j\le n} \log \left(1 + e^{\alpha + M_{ij}}\right) \nonumber \\
&& + \frac{1}{2\lambda}\left[\alpha - (z^m_\alpha - u^m_\alpha)\right]^2
+ \frac{1}{2\lambda}\|M - (z^m_M - u^m_M)\|^2_F, \label{eq:admm02} \\
x^{m+1}_L &=& \mbox{arg min}_{ L} \quad \delta \|L\|_\ast
+ \frac{1}{2\lambda}\|L - (z^m_L - u^m_L)\|^2_F,\label{eq:admm03}
\\
&& \mbox{subject to $L$ is positive semidefinite;} \nonumber \\
x^{m+1}_S &=& \mbox{arg min}_{ S} \quad  \gamma \|S\|_1
+ \frac{1}{2\lambda}\|S - (z^m_S - u^m_S)\|^2_F,
\label{eq:admm04}
\\
&& \mbox{subject to $S$ is symmetric,} \nonumber
\end{eqnarray}
where $\|\cdot\|_F$ is the matrix Frobenius norm, defined as
$\|M\|^2_F = \sum_{i,j} m^2_{ij}$ for a matrix $M = (m_{ij})$.
The problem in \eqref{eq:admm02} may not have a closed-form solution. We use a simple gradient descent to solve this step, setting the step size, $\alpha=0.1$ and stopping criteria, $\max(|x_{\alpha,m}^{(t+1)}-x_{\alpha,m}^{(t)}|,\|x_{M,m}^{(t+1)}-x_{M,m}^{(t)}\|_{\infty}) \leq 10^{-9}$. Note that there are close-form solutions to \eqref{eq:admm03} and \eqref{eq:admm04}, while \eqref{eq:admm02} is a unconstrained convex optimization problem. More specifically, in \eqref{eq:admm03}, suppose the eigenvalue decomposition of the symmetric matrix $(z^m_L - u^m_L)$ can be written as
$$
z^m_L - u^m_L = T \Lambda T^T,
$$
where $T$ is orthogonal ($T T^T = I_n$), then, for $J=I_{n}-\frac{1}{n}\mathbbm{1}\mathbbm{1}^T$, we have
$$
x^{m+1}_L = J (T \mbox{diag}(\Lambda-\lambda \delta)_+ T^T)J^{T},
$$
and diag$(\Lambda-\lambda \delta)_+$ is a diagonal matrix with the $j$th
diagonal entry being
$$
(\Lambda_{jj}-\lambda \delta)_+ = \left\{
\begin{array}{ll}
0, & \mbox{ if } \Lambda_{jj} < \lambda \delta \\
\Lambda_{jj}-\lambda \delta, & \mbox{ if } \Lambda_{jj} \ge \lambda \delta .
\end{array}
\right.
$$
In \eqref{eq:admm04}, we have, for $i \neq j$,
\[
S_{ij} = \left\{
\begin{array}{ll}
0, & \mbox{ if } |(z^m_S - u^m_S)_{ij}| < \lambda \gamma  \\
(z^m_S - u^m_S)_{ij}-\lambda\gamma , &
\mbox{ if } (z^m_S - u^m_S)_{ij} > \lambda \gamma \\
(z^m_S - u^m_S)_{ij}+\lambda\gamma , &
\mbox{ if } (z^m_S - u^m_S)_{ij} <- \lambda \gamma
\end{array}
\right.
\]
\\

@@

\noindent \textbf{Ad-hoc links} Ad-hoc links of synthetic network data can be thought of as $"$across edges$"$ between clusters of nodes. Since indices of non-zero entries of $S$ are randomly chosen, there might be some indices of non-zero $S_{ij}$ entries where $L_{ij}$ is also non-zero. In this case, this makes the edges indistinguishable if they come from $S_{ij}$ or $L_{ij}$. Some of "across

\noindent edges" are also attributed to $\alpha$, and unfortunately, our algorithm cannot make a difference whether the edge comes from $\alpha$ or $S_{ij}$. So the total number of across edges might not be exactly same as we set in our data generation setting. \\

@@

\begin{figure}[!t]
\includegraphics[width=1\textwidth]{Fig3.eps}
\caption{Illustration of first two eigenvectors of estimated $\hat{L}$ matrix for original
citation network with 543 papers(left) and sub-network graph(right). Presentation
on clustering results of k-means algorithm with different colors.}
\label{fig:figure1}
\end{figure}

@@

\[
h(\hat{\Theta})-h(\Theta^{*}) \geq 
\langle\, \nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
+ \frac{\tau}{2}\|\Delta_{\hat{\Theta}}\|_{F}^{2}
\]

Hence, we can rearrange the term in (\ref{eq:2}) as follows:

\begin{equation}\label{eq:3}
    \frac{\tau}{2}\|\hat{\Delta}^{\Theta}\|_{F}^{2} \leq 
    -\langle\,\nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
    + \delta \|L^{*}\|_\ast + \gamma \|S^{*}\|_1 
    - \delta \|\hat{L}\|_\ast - \gamma \|\hat{S}\|_1
\end{equation}

Here, we introduce another notation, for any pair of positive tuning parameters $(\delta,\gamma)$ which is defined as the weighted combination of the two regularizers : 
\[
\mathbb{Q}(L,S)   := \|L\|_{*} + \frac{\gamma}{\delta}\|S\|_1
\]
Through the definition of $\mathbb{Q}$, we can rewrite (\ref{eq:3}) as follows:
\begin{equation}\label{eq:4}
    \frac{\tau}{2}\|\hat{\Delta}^{\Theta}\|_{F}^{2} \leq 
    -\langle\,\nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
    + \delta \mathbb{Q}(L^{*},S^{*}) - \delta \mathbb{Q}(\hat{\Delta}^L + L^{*},\hat{\Delta}^S + S^{*})
\end{equation}

>>>>

\author[2]{Eric Heim}

\author[2]{Timothy Van Slyke}

\author[2]{Lee Seversky}

@@

@@

\section{Introduction}
\label{sec:intro}
We study a citation network, where each node (i.e., item) can be a technical report or a publication. 
A node may cite another node. 
Associated with a pair of nodes $i$ and $j$, we denote a binary random variable $X_{ij}$, where $1 \le i,j \le n$ and $n$ is the total number of nodes. 
We have $X_{ij}=1$ if and only if either node $i$ cites node $j$ or vice versa; otherwise $X_{ij}=0$. 
For each node $i$, we assume that there is an associated binary vector $f_{i} \in \mathbb{R}^K$, such that the $k$th entry of $f_{i}$, $f_{ik} = 1$, if and only if node $i$ is related to topic (i.e., factor) $k, 1\le k \le K$. 
Here $K$ is the total number of underlying topics (i.e., factors, or trends). 
We assume a logistic model for $X_{ij}$'s: for $1\le i,j \le n$,

\begin{equation}
\label{eq:logistic01}
\mathbb{P}(X_{ij}=1) = \frac{e^{\alpha + f_i^T D f_j }}{1 + e^{\alpha + f_i^T D f_j }},
\end{equation}
where $\alpha \in \mathbb{R}$ is a parameter and matrix $D \in \mathbb{R}^{K \times K}$ is a diagonal matrix: $D = \mbox{diag}\{d_{1},d_{2},\ldots,d_{K}\}$. 
We assume $d_{ii} > 0$ for $1 \le i \le K$.
Another way to put \eqref{eq:logistic01} is
\begin{equation}
\label{eq:logistic02}
\mathbb{P}(X_{ij}=1) = \frac{\exp\left\{\alpha + \sum_{k=1}^K f_{ik}f_{jk}d_{k} \right\}}{1 + \exp\left\{\alpha + \sum_{k=1}^K f_{ik}f_{jk}d_{k} \right\}}. 
\end{equation}
A justification of the above model is that when both node $i$ and node $j$ are related to topic $k$, they have a higher chance to cite one way or the other. 
We have assumed a common strengthen coefficient $d_k$ ($1\le k \le K$) for factor $k$, despite different nodes. 
We denote a matrix $F = \{f_1, f_2, \ldots, f_n\} \in \mathbb{R}^{K \times n}$. 
Each column $i$ in matrix $F$ contains the factor loadings associated with the node $i$ ($1\le i \le n$). 
Given the diagonal matrix $D$ and the factor loading matrix $F$, we assume that $X_{ij}$'s are independent; therefore we have the total conditional probability function as follows:
\begin{equation}
\label{eq:logistic03}
\mathbb{P}(\{X_{ij}, 1\le i,j \le n\})
= \prod_{1\le i<j \le n} \mathbb{P}(X_{ij})
= \prod_{1\le i<j \le n}  \frac{e^{X_{ij}(\alpha + f_i^T D f_j) }}{1 + e^{\alpha + f_i^T D f_j }},
\end{equation}
where $\mathbb{P}(X_{ij})$ is given in \eqref{eq:logistic02}.
The last equation holds because $X_{ij}$ only takes binary (i.e., $0$ or $1$) values.
Recall that the dot product of two matrices with same dimensionality, $A,B\in \mathbb{R}^{a \times b}$, is defined as $A\bullet B=\mbox{trace}(A^T B) = \sum_{i=1}^a\sum_{j=1}^b a_{ij}b_{ij}$.
The above \eqref{eq:logistic03} can be further rewritten as
\begin{equation}
\label{eq:logistic04}
\mathbb{P}(\{X_{ij}, 1\le i,j \le n\})
= \frac{\exp(\alpha \sum_{1\le i< j\le n}X_{ij} +\frac{1}{2} X \bullet (F^T D F))}{\prod_{1\le i<j \le n}  1 + e^{f_i^T D f_j }},
\end{equation}
where we assume $X_{ii}=0$ for all $i$ ($1\le i \le n$) and $X_{ij} = X_{ji}$ for all $i$ and $j$ ($1\le i,j \le n$), i.e., the matrix $X$ is symmetric.
The above delivers a factor analysis model. 
Various linear and nonlinear latent variable models have been studied extensively in the literature (e.g.,\cite{joreskog1969general, mcdonald2014factor,lord2008statistical, rasch1980probabilistic, harman1960modern, joreskog1970general}).

@@

We now present each of the three steps of the ADMM algorithm and show that the proximal
operators $\mathbf{P}_{\lambda,f}$ and $\mathbf{P}_{\lambda,g}$ are easy to evaluate.
Let
$$
x^m = (x^m_\alpha, x^m_M, x^m_L, x^m_S), \quad
z^m = (z^m_\alpha, z^m_M, z^m_L, z^m_S), \quad
u^m = (u^m_\alpha, u^m_M, u^m_L, u^m_S).
$$
Step 1. We solve $x^{m+1} = \mathbf{P}_{\lambda,f}(z^m - u^m)$.
Due to the special structure of $f(\cdot)$,
$x^{m+1}_\alpha, x^{m+1}_M, x^{m+1}_L$, and $x^{m+1}_S$
can be updated separately. 
More precisely, we have
\begin{eqnarray}
x^{m+1}_\alpha, x^{m+1}_M &=& \mbox{arg min}_{\alpha, M} \quad
-\frac{\alpha}{n} \sum_{1\le i< j\le n}X_{ij}
-\frac{1}{2n} X \bullet M
+ \frac{1}{n} \sum_{1\le i<j\le n} \log \left(1 + e^{\alpha + M_{ij}}\right) \nonumber \\
&& + \frac{1}{2\lambda}\left[\alpha - (z^m_\alpha - u^m_\alpha)\right]^2
+ \frac{1}{2\lambda}\|M - (z^m_M - u^m_M)\|^2_F, \label{eq:admm02} \\
x^{m+1}_L &=& \mbox{arg min}_{ L} \quad \delta \|L\|_\ast
+ \frac{1}{2\lambda}\|L - (z^m_L - u^m_L)\|^2_F,\label{eq:admm03}
\\
&& \mbox{subject to $L$ is positive semidefinite;} \nonumber \\
x^{m+1}_S &=& \mbox{arg min}_{ S} \quad  \gamma \|S\|_1
+ \frac{1}{2\lambda}\|S - (z^m_S - u^m_S)\|^2_F,
\label{eq:admm04}
\\
&& \mbox{subject to $S$ is symmetric,} \nonumber
\end{eqnarray}
where $\|\cdot\|_F$ is the matrix Frobenius norm, defined as
$\|M\|^2_F = \sum_{i,j} m^2_{ij}$ for a matrix $M = (m_{ij})$.
The problem in \eqref{eq:admm02} may not have a closed-form solution. 
We use a simple gradient descent to solve this step, setting the step size, $\alpha=0.1$ and stopping criteria, $\max(|x_{\alpha,m}^{(t+1)}-x_{\alpha,m}^{(t)}|,\|x_{M,m}^{(t+1)}-x_{M,m}^{(t)}\|_{\infty}) \leq 10^{-9}$. 
Note that there are close-form solutions to \eqref{eq:admm03} and \eqref{eq:admm04}, while \eqref{eq:admm02} is a unconstrained convex optimization problem. 
More specifically, in \eqref{eq:admm03}, suppose the eigenvalue decomposition of the symmetric matrix $(z^m_L - u^m_L)$ can be written as
$$
z^m_L - u^m_L = T \Lambda T^T,
$$
where $T$ is orthogonal ($T T^T = I_n$), then, for $J=I_{n}-\frac{1}{n}\mathbbm{1}\mathbbm{1}^T$, we have
$$
x^{m+1}_L = J (T \mbox{diag}(\Lambda-\lambda \delta)_+ T^T)J^{T},
$$
and diag$(\Lambda-\lambda \delta)_+$ is a diagonal matrix with the $j$th
diagonal entry being
$$
(\Lambda_{jj}-\lambda \delta)_+ = \left\{
\begin{array}{ll}
0, & \mbox{ if } \Lambda_{jj} < \lambda \delta \\
\Lambda_{jj}-\lambda \delta, & \mbox{ if } \Lambda_{jj} \ge \lambda \delta .
\end{array}
\right.
$$
In \eqref{eq:admm04}, we have, for $i \neq j$,
\[
S_{ij} = \left\{
\begin{array}{ll}
0, & \mbox{ if } |(z^m_S - u^m_S)_{ij}| < \lambda \gamma  \\
(z^m_S - u^m_S)_{ij}-\lambda\gamma , &
\mbox{ if } (z^m_S - u^m_S)_{ij} > \lambda \gamma \\
(z^m_S - u^m_S)_{ij}+\lambda\gamma , &
\mbox{ if } (z^m_S - u^m_S)_{ij} <- \lambda \gamma
\end{array}
\right.
\]

@@

\noindent \textbf{Ad-hoc links} Ad-hoc links of synthetic network data can be thought of as $"$across edges$"$ between clusters of nodes. 
Since indices of non-zero entries of $S$ are randomly chosen, there might be some indices of non-zero $S_{ij}$ entries where $L_{ij}$ is also non-zero. 
In this case, this makes the edges indistinguishable if they come from $S_{ij}$ or $L_{ij}$. 
Some of ``across edges" are also attributed to $\alpha$, and unfortunately, our algorithm cannot make a difference whether the edge comes from $\alpha$ or $S_{ij}$. 
So the total number of across edges might not be exactly same as we set in our data generation setting. \\

@@

\begin{figure}[!t]
\includegraphics[width=1\textwidth]{Fig3.eps}
\caption{Illustration of first two eigenvectors of estimated $\hat{L}$ matrix for original
citation network with 543 papers(left) and sub-network graph(right). Presentation
on clustering results of k-means algorithm with different colors.}
\label{fig:figure3}
\end{figure}

@@

\begin{equation}
\label{eq:2}
h(\hat{\Theta})-h(\Theta^{*}) \geq
\langle\, \nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
+ \frac{\tau}{2}\|\Delta_{\hat{\Theta}}\|_{F}^{2}
\end{equation}
Consequently, we can rearrange the term in (\ref{eq:2}) as follows:
\begin{equation}\label{eq:3}
    \frac{\tau}{2}\|\hat{\Delta}^{\Theta}\|_{F}^{2} \leq
    -\langle\,\nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
    + \delta \|L^{*}\|_\ast + \gamma \|S^{*}\|_1
    - \delta \|\hat{L}\|_\ast - \gamma \|\hat{S}\|_1
\end{equation}

Here we introduce another notation, for any pair of positive tuning parameters $(\delta,\gamma)$ which is defined as the weighted combination of the two regularizers :
\[
\mathbb{Q}(L,S)   := \|L\|_{*} + \frac{\gamma}{\delta}\|S\|_1
\]
Through the definition of $\mathbb{Q}$, we can rewrite (\ref{eq:3}) as follows:
\begin{equation}\label{eq:4}
    \frac{\tau}{2}\|\hat{\Delta}^{\Theta}\|_{F}^{2} \leq
    -\langle\,\nabla_{\Theta}h(\Theta^{*}),\hat{\Theta}-\Theta^{*}\rangle\
    + \delta \mathbb{Q}(L^{*},S^{*}) - \delta \mathbb{Q}(\hat{\Delta}^L + L^{*},\hat{\Delta}^S + S^{*})
\end{equation}
